Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-16 write-ups/PLDI/paper.tex
Space.

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.

Workgroups of a \emph{cooperative} kernel \emph{are} fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\end{figure}

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. For multi-kernel
kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
primitives via a per kernel \emph{context}.  For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
 It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation.

@@

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per global barrier call and, depending on
the order in which workgroups arrive at the barrier, it is possible
that only one workgroup is killed per barrier call.  This may not
allow the scheduler to gather workgroups required to execute another
task in a timely fashion.

>>>>

Because workgroups of a cooperative kernel are fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\end{figure}

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. For multi-kernel
kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to other primitives.  It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation.

@@

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per barrier call and, depending on
order of arrival, it is possible
that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.
